Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-01 manuscript.tex
ALF tex updates, bib updates, manuscript_annotatedALF2.pdf
Updates from Overleaf

%include hyperlinks to figures
\documentclass{article}
\usepackage{graphicx}
\usepackage[%  
    colorlinks=false,
    pdfborder={0 0 0},
    linkcolor=blue
]{hyperref}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{CJK}
\usepackage{framed}
\usepackage[hidelinks]{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

@@

%\author{I. Ken Groupleader}
%\altaffiliation{A shared footnote}
%\email{i.k.groupleader@unknown.uu}
%\phone{+123 (0)123 4445556}
%\fax{+123 (0)123 4445557}
%\affiliation[Unknown University]
%{Department of Chemistry, Unknown University, Unknown Town}
%\alsoaffiliation[Second University]
%{Department of Chemistry, Second University, Nearby Town}

%\author{Susanne K. Laborator}
%\email{s.k.laborator@bigpharma.co}
%\affiliation[BigPharma]
%{Lead Discovery, BigPharma, Big Town, USA}

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}. A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}. Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}. Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspects the overall dynamics process in one direction or in a biased fashion.

@@

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system  -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that G:C base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that higher resolution MSMs (measured by a reduction in the required lag time) are accessible. Additionally, we compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs. Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

@@

The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. 3spn2 been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.774 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat to account for implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to maximize transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. Depending on sequence, we observed between 55-100 hybridizing and dissociating transitions across all runs.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs to estimate the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models. (Can include equations or additional info in SI?) We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordiantes (figure \ref{fig:allseq_features_vamp2}). We found that reciprocal coordinates performed better across the board, but we saw no substantial difference between the three featurization methods. This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model. The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution. These features were normalized and passed into sequence-specific SRVs.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=80mm,
        scale=0.5]{Figs/figs_imp/all_permute_vamps_nolegend.png}
        \caption{5-fold cross validation to calculate the VAMP-2 score for each feature set. The inverse distances showed improvement across sequences, and the 55-dimension symmetrized coordinate set performed about as well as larger features set.}
        \label{fig:allseq_features_vamp2}
	\end{center}
\end{figure}

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

@@

MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results and supplemental information.

@@

% Include analysis on overall hybridization rates here or at the end?
% Show difference between between full hybridization and partial events (GC-end and AT-all)

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=400, scale=1]{Figs/figs_imp/AT-all_implied_timescales.png}
        \caption{SRV-MSM implied timescales converged faster than tICA-MSM implied across the five leading AT-all modes. Timescales are directly compared at the chose lag time of 1.2 ns}
        \label{fig:AT-all_dynamic}
	\end{center}
\end{figure}

%addition to timescale comparison, we found that tICA-MSMs often produced inconsistent PCCA+ clusterings between runs and tended to be less physically meaningfully

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Figs/figs_imp/allseq_states.png}
        \caption{Nearest neighbor representations, molecular renderings, and sequence-dependent probabilities for PCCA+ macrostates. The dissociated state is not shown but represents the remainder of the stationary probability for each sequence.}
        \label{fig:allseq_table}
	\end{center}
\end{figure}

\begin{figure}[hb!]
	\begin{center}
        \includegraphics[width=160mm, 
        scale=1]{Figs/figs_imp/allseq_transitions.png}
        \caption{Free energies maps and PCCA+ states clusterings are shown in the tICA space in order to enhance visibility in dimensions. Flux diagram between PCCA+ states are shown for each sequences, accompanied by representative structures are for each state. Arrows indicate the probability of transitioning out of the state within the lag time. Circle areas are proportional to state stability.}
        \label{fig:allseq_transition}
	\end{center}
\end{figure}

\subsection{Interpreting PCCA+ Macrostates}

Following the analysis pipeline described above, we generate SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. These include the fully hybridized state (H) in which all native base pairings are intact, the fully dissociated state (D), four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, and a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound. Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5' vs. 3') and the number of shifted motifs relative the native state (1 vs. 2). GC-mix displays two-state behavior within the resolution of the model, however we will show that the hybridization and dissociation are still characterized by an ensemble of shorter-lived states. Stationary probabilities in each of these states are shown in figure \ref{fig:allseq_table}. We also show that PCCA+ clustering assigns states to relative free energy minima in the tICA space.

% rewrite this into one paragraph
\subsection{Shifted state stability}

>>>>

%include hyperlinks to figures
%\documentclass{article}
\usepackage{graphicx}
\usepackage[%  
    colorlinks=false,
    pdfborder={0 0 0},
    linkcolor=blue
]{hyperref}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{CJK}
\usepackage{framed}
%\usepackage[hidelinks]{hyperref}
%\usepackage{algorithm}
%\usepackage{algpseudocode}
%\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

@@

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}. A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}. Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}. Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if long-lived metastable states facilitate the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that rapidly perturb DNA and monitor relaxation to a new equilibrium \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to provide molecular level detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied the hybridization process as a relaxation to a chosen equilibrium or in a biased fashion.

@@

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to perform unbiased simulations of hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We synthesize the simulation trajectories using Markov State Models (MSMs) to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and implement state free-reversible VAMPnets (SRVs) within the MSM parameterization pipeline to construct higher resolution kinetic models than are achievable using standard time-lagged independent components analysis (tICA) based approaches.

% this section should look more like a conclusion foreshadowing conclusions and main take aways 
We find that 3spn2 allows us to access long length and time scales and observe sequence-dependent kinetics and relaxation rates that are, within a corrective scaling factor, in agreement with observed experimental trends. Applying SRV-MSMs enables the construction of high-resolution sequence-dependent kinetic models of hybridization and dissociation. Consistent with previous studies \citep{Phys2014,Romano2013DNADependence,Araque2016LatticeCooperativity}, we find the degree of repetitiveness in the sequence and therefore the kinetic accessibility and thermodynamic stability of shifted states leads to richer folding dynamics populated by a diversity of long-lived metastable states. Disrupting repetitive stretches of bases by judicious placement of interrupting G:C pairs can tune the landscape from rich 6-state to very simple 2-state (“all-or-nothing”) behavior. Furthermore, concentrating G:C pairs at the center of a duplex produces a longer-live frayed state that facilitates the hybridization and dissociation process. Analysis within metastable state populations using diffusion maps allows us to explore the conformational diversity and heterogeneity within each state at time scales shorter than the kinetic resolution of the SRV-MSMs. This provides us with molecular-level understanding for the differential stability of various conformational states as a function of oligomer sequence. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

% example text from Andy's conclusions:

% Correlation of the data-driven kinetic order parameters extracted by SRVs provides new understanding of the sequence-dependent kinetic mechanisms by which oligomers hybridize and reveal these to be richer than classical base-pair shifting and “all-or-nothing” behaviors.

% original last paragraph
%By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that higher resolution MSMs (measured by a reduction in the required lag time) are accessible. Additionally, we compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs. Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

% introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

@@

The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interaction sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. The 3spn2 model been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.8 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble. Temperature was controlled by a Langevin thermostat that models interactions with an implicit water solvent. \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to observe frequent transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequence, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. We observed between 55-100 hybridization/dissociation events for each of the four sequences.

@@

A featurization of trajectories was required to capture fundamental dynamics while accounting for translational and rotational symmetries. Pairwise distances between base centers of mass -- similar to heavy atom distances in protein models \citep{Sidky}-- satisfied these requirements and were calculated at each frame using the MDtraj Python library \citep{McGibbon2015MDTraj:Trajectories}. Although pairwise distances remove translational and rotational symmetries, additional symmetries arise from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together into a "symmetrized" feature set following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. Additional details on the VAMP-2 score can be found in the supplemental information. By comparing scores across several candidate feature sets (figure \ref{fig:allseq_features_vamp2}), we found that minimal kinetic information was lost in the  symmetrization procedure. Furthermore, the smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=80mm,
        scale=0.5]{Figs/figs_imp/all_permute_vamps_nolegend.png}
        \caption{A comparison of VAMP-2 scores via 5-fold cross-validation for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordinates. Reciprocal coordinates performed better across the board, but there was no substantial difference between the three featurization methods. This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model.}
        \label{fig:allseq_features_vamp2}
	\end{center}
\end{figure}

@@

The next step in the MSM pipeline is to create a low-dimensional kinetic projection where state clustering is performed. While most pipelines employ tICA to generate a linear transformation of input features, we use SRVS as a more powerful modular replacement for tICA. SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been used to construct high kinetic resolution MSMs applied to molecular simulations of WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}.

@@

% need some more detail here -- talk to Andy about this
MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. After Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results and supplemental information.

@@

% Include analysis on overall hybridization rates here or at the end?

\subsection{Sequence-dependent coarse-grained kinetics agree with T-jump IR data}

% rerun these at the actual final temps and just see what happens? Doesn't really matter where Tjump occurs and could make things easier as a whole...
% talk about rerunning AT-all and GC-mix=

As an initial validation of the 3spn2 kinetics, we compared simulated relaxation measurements with temperature-jump (T-jump) experiments. Coarse-grained timescales are inherently difficult to match against experiments due to smoothing of the free energy landscape and variable acceleration across different degrees of freedom. As such, we evaluated multiple temperatures for each sequence and inferred acceleration factors from temperature-dependent trends.  Experimental temperature-dependent relaxation timescales were measured for each sequence using T-jump IR spectroscopy as described previously \citep{Sanstead2018DirectDehybridization}. We recorded "fast" and "slow" amplitude-weighted responses -- previously attributed to terminal base pair fraying and duplex dissociation, respectively -- for each temperature and sequence. To compare results with 3spn2, we ran 120 shorter (1 us) simulations initialized in the hybridized state at final T-jump temperatures for each sequence. We derived relaxation times for a slow dissociation response and fast fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits.

% need to discuss tmelt shift
We measured the slow dissociation response by fitting the distribution of times at which the core base pairs separate beyond a 2 nm cutoff. We found the inverse of these relaxation times -- the effective dissociation rate -- to increase exponentially with temperature, which is expected given the large enthalpic barrier of dissociation. We noted an average temperature shift of about 4C between experiment and simulation exponentials. While 3spn2 captures melting temperature relatively well, some deviation is expected given varied ion conditions and other coarse-grained effects. Furthermore, we see an acceleration of about one order of magnitude compared to experiment, although this factor is again sensitive to the exact definition of melting temperature. After accounting for the melting temperature shift and acceleration factor, we saw agreement between the experimental data and simulated relaxation fits. GC-core showed the largest deviation experiment, indicating that increased temperature triggers dissociation at a disproportionately higher rate in simulation compared to experiment.

% GC-core fast timescales could be slower as they have contributions from internal fraying which would have a longer relaxation than the termini alone.
% in order to make comparisons across these different free energy landscapes, we adjusted simulations by a consistent acceleration factor. We maintained a consistent accegqrelation factor across sequences, but found that optical scaling different by about one order of magnitude between fast and slow responses. It is expected given that different degrees of freedom often experience differing amounts of accelerations in coarse-grained models.

% need to update this fig with labels and temp shift

\begin{figure}  %[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/Tjump_responses.png}
        \caption{Fitting temperature-dependent trends to the "slow" dissociation mode and "fast" fraying mode detected in experiments. The effective simulation temperature is shifted by 4 degrees celcius to account for systematic differences in simulation and experimental melting temperature. Different scaling factors are applied to fast and response to account for variable coarse-grained acceleration along different degrees of freedom.}
        \label{fig:relaxation-comparison}
	\end{center}
\end{figure}

The fast response -- which Sanstead et. al attributed to terminal base pair fraying signatures -- was more challenging to compare against simulation observables. Numerous experimental and computational studies have shown that DNA and RNA fraying is a complex dynamical process with timescales that span 5 ps to several microseconds \citep{ Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix, Galindo-Murillo2015ConvergenceDGCACGAACGAACGAACGC}. All-atom simulations suggest that frayed ends can assume misaligned WC bonds, base-sugar hydrogen bonds, and terminal stacked conformations \citep{PinamontiTheModels, Zgarbova2014BaseRNA}. Given that there is only one interaction site parameterized on each 3spn2 base, we would not expect to resolve this diverse collection of states and dynamics. Instead, we measured the fast fraying response by counting frames until either duplex terminal end splits beyond a cutoff. This approach assumes that fraying on the permutable top and bottom of the duplex are independent from each other, and that a base pair distance is a reliable approximation for the ensemble spectroscopy signal. This is a reasonable assumption given that the amplitude-weighted timescales should consist largely of terminal fraying events. Again we fit relaxation curves to the ensemble of fraying timescales in order to extract rate approximations for each sequence at a series of temperatures.
